---
title: "Intro to R"
subtitle: "Day 7"
date: "Class date: 02/28/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/06/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

```{r echo=FALSE}
library(knitr)
read_chunk('analysis.R')
```




# Libraries

```{r, 'libs', echo=TRUE, message=FALSE}
```

```{r, 'setwd', echo=FALSE}
```

# Load data

We have a directory with 5 different .csv data files. We need 
to use a different method to load all 5 files and combine them 
into a single dataframe. 

```{r, 'loadData', echo=TRUE, warning=FALSE}
```

# Tidy data

The ```f0``` column has many undefined data points (and is set 
as a factor). We need to change these values to ```NA``` and 
set the column as numeric. We also need to create a few new 
columns. We will use the ```separate``` function from the 
```tidyr``` package.

```{r, 'cleandata', echo=TRUE}
```

# Plots

Let's visualize what we did. 

```{r, 'densityVOT', echo=TRUE, fig.width=10, fig.height=5, fig.retina=2, fig.align='center'}
```

```{r, 'densityVOTindividual', echo=TRUE, fig.width=10, fig.height=5, fig.retina=2, fig.align='center'}
```

```{r, 'vowelPlot', echo=TRUE, fig.width=10, fig.height=5, fig.retina=2, fig.align='center'}
```

</br>

You can download the data [here](http://www.jvcasillas.com/585_01_s2017/programming/r/07/data.zip)

</br></br></br></br>



