---
title: "Intro to R"
subtitle: "Day 2"
date: "Class date: 01/24/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/02/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

# knitr

## .R scripts

We can create a reproducible document from a .R script using `spin()`

## .Rmd files

When we want to generate a report that will include prose, an `RMarkdown` 
document (.Rmd) is probably better. 

### yaml front matter 

Some important details about our document are contained in the yaml front 
matter. It is the section at the top of the document that looks like this:

```
---
title: 'title here'
output: html_document
---
```

We can change the type of document to .doc (`output: word_document`) or 
.pdf (`output: pdf_document`). 

### Code chunks

If we want to include R code in our .Rmd file, we must use code chunks. 
An R code chunk begins with 3 backticks \`\`\`, curly braces containing 
the letter "r", followed by our code. The chunk is closed with 3 more 
backticks. Ex. 

<pre><code>```{r}
2 + 2
```</code></pre>

### Inline code 

If we want to evaluate an R expression inside of prose, we can use the 
following syntax: 

```{r, eval=FALSE}
    Inline code looks like this `r 2 + 2`
```

This is an example. Two plus two equals `r 2 + 2`. 



</br>

# Data structures (cont.)

## Assignment

We can assign values to objects using the assignment operator `<-`. 

```{r}
# Assign value 5 to 'object'
object <- 5

# print 'object' 
print(object)

# Add 5 to 'object'
object + 5
```

## Vectors (cont.)

The most basic data structure is the vector. Vectors have types (numeric, character, etc.). 
If we want to create a vector with a length larger than 1, we can use the `c()` function 
(concatenate). 

```{r}
# assigning variables 
x <- c(1, 2, 6, 3, 7)
y <- c("a", "b", "c", "d", "e")

# print them 
print(x)
print(y)
```

Vectorized math works the same as an object of length 1 (compare with `object + 5` above). 

```{r}
x + 20
```

## Dataframes

A dataframe is a two dimensional object made of columns and rows. The columns have names 
and can be of different types; however, all columns must have the same length. We can 
create a dataframe using the `data.frame()` function. 

```{r}
# creating dataframes
df <- data.frame(x, y)

print(df)
```

We can get information about a dataframe using `str()`, `head()`, `tail()`, and `dim()`. 

```{r}
# Exploring objects 
str(df)
head(cars)
tail(cars)
dim(cars)
```

## Lists

A list is a data structure that can cantain any type of object. 

```{r}
myList <- list(a = 1:3, b = 'abc', c = TRUE)
str(myList)
```

You will probably not use lists too much, but it is a good idea to understand how 
they work in order to have a better understanding of R syntax. 





</br>

# Subsetting

## Subsetting vectors 

We can subset a vector using brackets (`[]`). Given that a vector is a 1D object, the number 
we include in the bracket refers to the position in the vector. In other words, if we want 
to see the first element of `x`. We can use the bracket notations `x[1]`.

```{r}
x[1]
```

## Subsetting dataframes

Subsetting a dataframe works in the same way; however, since the structure has two dimensions, 
it is necessary to give information about both. We will still use the bracket notations, but we 
will designate `rows` and `columns` by separating them with a comma. 

```{r}
# First row, all columns
df[1, ]

# Secont row, first column 
df[2, 1]

# rows 1 to 3, all columns
df[1:3, ]
```

We can also subset using the names of the columns:

```{r}
# Give me the first 10 values of 
# 'speed' in the 'cars' data frame 
cars$speed[1:10]

# another way 
cars[1:10, 'speed']
```



## Subsetting lists

We can single out an element of a list using bracket notation (like with a vector). This will 
return another list. 

```{r}
# Give me the first element
myList[1]

# Show me the structure of the first element
str(myList[1])
```

This may not always be useful. For example, is the first element is a vector, we may want it 
as a vector (and not a list). In order to subset an element of a list and get it in its `natural` 
state, we can use double brackets (Note: this is useful if you need to perform an operation on 
something in a list).

```{r}
# Give me the first element (not as a list)
myList[[1]]

# Show me the structure 
str(myList[[1]])
```

We can also refer to the names of the objects inside the list directly (if they have names).

```{r}
myList['a']
myList[['a']]
```

We can also subset a subset of a list (what?). For example, the first element of `myList` is 
the vector `a`. We can obtain the second values of `a` like this:

```{r}
myList[['a']][2]
```





</br>

# Descriptive statistics

We can calculate descriptive statistics on vectors. Some handy functions are: 

- `sum()`
- `mean()`
- `sd()`
- `abs()`

```{r}
# Mean of 'x'
mean(x)

# Sum of 'x'
sum(x)
```

We can also do this using data frames or lists. 

```{r}
# Mean of 'x' in dataframe 'df'
mean(df$x)

# SD of 'x' in dataframe 'df'
sd(df$x)

# Sum of 'a' is list 'myList' (double brackets are obligatory)
sum(myList[['a']])

# SD 'speed' in 'cars'
sd(cars$speed)
sd(cars[, 'speed'])

# Mean of first 15 rows of the first column 
mean(cars[1:15, 1])
mean(cars$speed[1:15])
mean(cars$speed[c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15)])
```

The `aggregate()` function is extremely handy. You provide a formula, 
the function you would like to apply, and the dataframe. For example, 
consider the `iris` dataset. 

```{r}
str(iris)
head(iris)
```

Let's pretend we are interested in the `Sepal.Length` (column 1). We 
notice that the `str()` function indicates that there is a factor 
called `Species` that has three levels ("setosa", "versicolor", "virginica").
Let's calculate the mean `Sepal.Length` for each level of `Species`. 
A simple way to do that is to use `aggregate()`. 

```{r}
# Mean Sepal.Length as a function of Species
aggregate(Sepal.Length ~ Species, FUN = mean, data = iris)
```




</br>

# Coercion 

We can change the class of an object through coercion. 

- as.data.frame()
- as.numeric(y)
- as.character(x)
- as.factor()

# Data sets

R is installed with a number of data sets that can be used to practice. To 
see them use `data()`. Here are the first 60...

```{r, echo=FALSE}
head(data()$results[, 3], 60)
```




</br>

# Basic bivariate plot

We can create a bivariate scatterplot using the function `biVarPlot()` from 
the package `lingStuff`.

```{r, fig.retina=2}
library(lingStuff)
biVarPlot(data = cars, x = dist, y = speed, family = 'Gaussian')
```

In base R we can do the same using the `plot()` function. 

```{r, fig.retina=2}
plot(speed ~ dist, data = cars)
```
