---
title: "Intro to R"
subtitle: "Day 4"
date: "Class date: 02/07/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/04/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```


# ggplot2

## Info 

- R package for making elegant plots quickly
- Written by Hadley Wickham when he was a graduate student at Iowa State
- Automatcally deals with spacings, text, titles but also allows you to annotate by “adding”
- Superficial similarity to lattice but generally easier/more intuitive
- Plots are built up in layers – Plot the data, overlay a summary, metadata and annotation
- Works with a data frame

## Components

- aestheic mappings: how data are mapped to color, size
- facets: for conditional plots
- geoms: geometric objects like points, lines, shapes
- stats: statitcal transformations
- scales: what scale an aesthetic map uses (example: male = red, female = blue)
- coordinate system

The basic structure is as follows...

```{r, eval=FALSE}
ggplot(dataframe, aes(x = , y = )) + 
```  

Aesthetics can include `x`, `y`, `shape`, `color`, `fill`, and others. 

You then add geoms (`geom_point()`, `geom_smooth()`, `geom_boxplot()`, among others).

```{r}
library(tidyverse)

ggplot(mtcars, aes(x = cyl, y = mpg)) + 
  geom_point()

ggplot(mtcars, aes(x = as.factor(cyl), y = mpg, fill = as.factor(vs))) + 
  geom_boxplot()

ggplot(mtcars, aes(x = hp, y = mpg, shape = as.factor(vs))) + 
  geom_point() 

ggplot(mtcars, aes(x = as.factor(cyl), y = mpg)) + 
  geom_boxplot() + 
  facet_grid(as.factor(vs) ~ .)
```

# Download data from the internet

We can download a file from the internet directly in R. 

```{r}
# Optionally assign the url to an object 
fileUrl <- 'https://dl.dropboxusercontent.com/u/7256527/data.csv'

# Download file 
download.file(fileUrl, destfile = "./data.csv", method = "curl")
```

# Data reformatting

## Wide to long

```{r}
set.seed(1)
wide <- data.frame(
  participant = c("p1", "p2", "p3", "p4", "p5", "p6"), 
  info = c("g1m", "g1m", "g1f", "g2m", "g2m", "g2m"),
  day1score = rnorm(n = 6, mean = 80, sd = 15), 
  day2score = rnorm(n = 6, mean = 88, sd = 8)
)

print(wide)
```

As you can see, this data set is in wide format. In order to do 
most analyses, we want **tidy data**. That is, each column should 
represent a variable and each row should represent an observation. 
A simple way to reformat the data is with the `gather()` function 
from the `tidyr` package. The basic format is as follows:

```{r eval=FALSE}
gather(df, key = groupingVar, val = dv)
```


```{r}
library(tidyr)

df_long <- gather(data = wide, key = day, val = score, -participant, -info)
print(df_long)

ggplot(df_long, aes(x = day, y = score)) + 
  geom_boxplot()
```

## More subsetting with logical statements

- `==`: equal 
- `!=`: not equal
- `>`: greater than
- `<`: less than
- `>=`: greater than or equal to
- `<=`: less than or equal to 
- `|`: or 
- `&`: and 

```{r}
# All rows in which `cyl` does not equal 6
mtcars[mtcars$cyl != 6, ]
```

## Add new variables 

We can create new variables in our data frame using logical statements. 
For example, the following code will create a new variable `size`. The 
variable is a factor with two levels: 'big' and 'small'. Any observation 
in which `cyl` size is 8 will be assigned 'big', any `cyl` size smaller 
(i.e. not equal to 8) will be assigned 'small'. 

```{r}
mtcars$size <- NA 
mtcars[mtcars$cyl == 8, 'size'] <- 'big'
mtcars[mtcars$cyl != 8, 'size'] <- 'small'
print(mtcars)
```



