---
title: "Intro to R"
subtitle: "Day 3"
date: "Class date: 01/31/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/03/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

# Load libraries

```{r}
library(tidyverse)
```

# Working directories and project folders

We are going to create a project folder `rsession` and set it as the working 
directory. In my case, the `rsession` folder is on the desktop. 

```{r, eval=FALSE}
# set working directory to 'sression'
setwd("~/Desktop/rsession")
```

# Loading data 

## From R

Remember R comes with data sets we can use to practice. 

```{r, echo=FALSE}
head(data()$results[, 3], 20)
```

## From an external source

We will download data from the internet and load it into R. Use the following 
[link](http://tinyurl.com/h6vnd8l) to get the data, save it as a .csv file and 
place it the `rsession` folder. (Note: I placed mine inside a subfolder called 
'data'. You can put yours wherever you want, but remember that you must specify 
the path to the file.)

In order to load the data we will use the `read.csv()` function. You can check 
which arguments this function takes by using the `args()` function (if you are 
curious). 

```{r}
df <- read.csv('./data/data.csv')
```

Now we can look at the data frame and obtain some summary information. 

```{r}
# Print data frame
print(df)

# 5 number summary
summary(df)

# structure
str(df)
```

Notice that this data frame contains two factors. We have not spent a lot of 
time dealing with factors. The two factors in this data frame are `group` and 
`gender`, which have 3 and 2 levels respectively. The `str()` function tell 
us all the information we need to know.



# Summary stats

The `aggregate` function can be useful when we want to calculate summary stats 
on factors. `aggregate` takes 3 important arguments: 1) the formula, 2) the FUN(ction), 
and 3) the data. Here are some examples:

```{r}
# read: mean score as a function of gender
aggregate(score ~ gender, FUN = sd, data = df)

# read: mean score as a function of group
aggregate(score ~ group, FUN = mean, data = df)

# read: mean score as a function of group and gender
aggregate(score ~ group + gender, FUN = mean, data = df)

# We can also calculate the mean and the sd at the 
# same time 
aggregate(score ~ gender, data = df, FUN = function(x) c(mean = mean(x), sd = sd(x)))
```


# More plots

Base R includes the `boxplot()` function, which can be useful for quick and 
simple data visualization. The argument structure is similar to `aggregate`. 
First we include the formula and then we specify the data frame.

```{r, fig.retina=2, fig.width=9, fig.height=5, fig.align='center'}
boxplot(score ~ gender + group, data = df)
```

# Projects

We can automate the project creation process using the `create_project()` function 
in the `lingStuff` package. 

</br></br></br></br>