## Install packages

### From CRAN

Packages can be installed from RStudio using the 'Packages' tab (normally in 
the bottom right window). Search for the package you want and hit `enter`. You 
can *manually* install packages from the console using the `install.packages()` 
function. 

```{r, 'install_cran', eval=FALSE, echo=TRUE}
install.packages('devtools')
```

### From GitHub

You can also install a package from GitHub. Often times these are packages in 
active development (before they are put on CRAN). In order to install a 
package from GitHub you need to know the name of the user and the name of the 
repo. For example, in this tutorial we will use the [lingStuff][lingStuff] 
package. It can be install using the following code: 

```{r, 'install_github', eval=FALSE, echo=TRUE}
library(devtools)
install_github("jvcasillas/lingStuff")
```

### From other sources

It is also possible to install packages from other sources (i.e., Bitbucket). 
See the ```devtools``` documentation for more information. 


