## Assigning variables

R uses objects. You must assign them using the assignment operator, ```<-```, 
in order to store them in memory. You can then call an object by typing its 
assigned name or using ```print()```. 


Store the value '2' to the object 'x'

```{r, assignment}
x <- 2
```

```{r, assignment1, exercise=TRUE}
x <- 2
print(x)
```

</br>

You can now perform an operation on 'x'.

```{r, assignment2, exercise=TRUE, exercise.setup = "assignment"}
x + 4
```

</br>

Try one of your own

```{r, assignment3, exercise=TRUE, exercise.setup = "assignment"}

```

```{r, assignment3-hint}
# Type in the variable 'x' and then add, subtract, multiple or divide it by 
# another number
```

If you assign another value to 'x', then the new value replaces the old one. 
Set 'x' to 30 and calculate the square root. 

```{r, assignment4, exercise=TRUE}

```

```{r, assignment4-hint-1}
# Remember that ```<-``` is the assignment operator  
# Use ```sqrt()``` to calculate the square root
```

```{r, assignment4-hint-2}
x <- 30
sqrt(x)
```

You can assign almost anything to an object. In the next example, assign some 
values to the objects I have given you and then use them to calculate a z-score.

```{r, assignment5, exercise=TRUE}
# enter values
my_score  <- 
class_avg <- 
class_sd  <- 

# calculate z-score (score - mean / sd)
my_z_score <- 

# print value
print(my_z_score)
```


```{r, assignment5-hint-1}
# enter values
my_score  <- 52.7
class_avg <- 48
class_sd  <- 10
```

```{r, assignment5-hint-2}
# enter values
my_score  <- 52.7
class_avg <- 48
class_sd  <- 10

# calculate z-score (score - mean / sd)
my_z_score <- (my_score - class_avg) / class_sd

# print value
print(my_z_score)
```




<!--

```{r math, echo=FALSE}
x <- 42
question(sprintf("Suppose $x = %s$. Choose the correct statement:", x),
  answer(sprintf("$\\sqrt{x} = %d$", x + 1)),
  answer(sprintf("$x ^ 2 = %d$", x^2), correct = TRUE),
  answer("$\\sin x = 1$")
)
```

### Quiz

```{r quiz1, echo=FALSE}
quiz(caption = "Quiz 1",
  question("What number is the letter A in the *English* alphabet?",
    answer("8"),
    answer("14"),
    answer("1", correct = TRUE),
    answer("23")
  ),
  question("Where are you right now? (select ALL that apply)",
    answer("Planet Earth", correct = TRUE),
    answer("Pluto"),
    answer("At a computing device", correct = TRUE),
    answer("In the Milky Way", correct = TRUE),
    incorrect = "Incorrect. You're on Earth, in the Milky Way, at a computer."
  )
)
```

-->
