
## Data structures

### Overview

There are four fundamental data structures used in R: 

- vectors
- matrices
- dataframes
- lists

### Vectors

The most important data structure is the **vector**. A vector is a series of 
numbers, integers, characters or logicals. The key is that all of the members 
of the vector have to be of the same type. 

```{r, vectors1, exercise=TRUE}

# 3 vectos: x, y, and z
x <- c(1, 2, 3)
y <- c(4, 5, 6)
z <- c('a', 'b', 'c')

# We can use the print function to see the vectors in the console
print(x)
print(y)
print(z)

```

We can perform operations on vectors. The following code will create a vector 
of numbers, assign it to the object ```my_vec``` and add one to each value in 
the vector (later we will see that this is much more important that it seems).

```{r, vectors2, exercise=TRUE}
# create vector 'my_vec'
my_vec <- c(10, 20, 30, 40, 50)

# Add one to each element of 'my_vec'
my_vec + 1 

```

If we have two vectors of the same class we can do vectorized math

```{r, vectors3, exercise=TRUE}
my_vec <- c(10, 20, 30, 40, 50)
y <- c(5, 10, 15, 30, 60)

# Add the values in y to the values in my_vec 
# think about the outcome before you do it

```

Create your own vector and perform an operation on it (note: this will only work 
if both vector are of the same type): 

```{r, vectors4, exercise=TRUE}
# Create numeric vector and assign to the object "x"

# Perform an operation on it (for example x^2)

# Create a character vector (call it 'y') and add the values of y to the values 
# of x. This gives an error. Why?

```


### Matrices 

A matrix is a 2d vector. You probably will not use them very often.

```{r, matrices, exercise=TRUE}
# Matrices = 2d vectors
matrix(1:10, nrow = 5, ncol = 2)

```



### Dataframes

Dataframes = 2d matrices with column and row names. ```IMPORTANT```: a column in 
a dataframe is like a vector, each column can be of a different type (factor, 
numeric, etc.). This is the closest equivalent to an excel sheet. You will 
probably spend most of your time in R working with dataframes.


```{r, dataframes1, exercise=TRUE}

x <- c(1, 2, 3)
y <- c(4, 5, 6)
z <- c('a', 'b', 'c')

# createa a dataframe using the vectors 'x', 'y', and 'z'
testDF <- data.frame(x, y, z)
print(testDF)

```

Let's look at some more interesting dataframes. The ```head()``` function shows 
the 1st six rows of a dataframe and the ```tail()``` function shows the final 
six rows. 

```{r, dataframes2, exercise=TRUE}
# Get first 6 rows
head(mtcars)

# Get last 6 rows
tail(USArrests)

```

We can get more information about a dataframe using the ```str()``` and 
```summary()``` functions. Try them below on the ```ToothGrowth``` dataset.

```{r, dataframes3, exercise=TRUE}


```
