
## Data structures

### Overview

There are four fundamental data structures used in R that we will discuss here: 

- vectors
- matrices
- dataframes
- lists

### Vectors

The most important data structure in R, in my opinion, is the **vector**. A 
vector is a series of numbers, integers, characters or logicals. The key is 
that all of the members of the vector have to be of the same type. 

```{r, vectors}
x <- c(1, 2, 3)
y <- c(4, 5, 6)
z <- c('a', 'b', 'c')
```

```{r, vectors1, exercise=TRUE}
# 3 vectos: x, y, and z
x <- c(1, 2, 3)
y <- c(4, 5, 6)
z <- c('a', 'b', 'c')

# We can use the print function to see the vectors in the console
print(x)
print(y)
print(z)

```

</br>

We can perform operations on vectors. 

```{r, vectors2, exercise=TRUE, exercise.setup = "vectors"}
x + 1 
```

</br>

If we have two vectors of the same class we can do vectorized math

```{r, vectors3, exercise=TRUE, exercise.setup = "vectors"}
x + y
```

</br>

Create your own vector and perform an operation on it (note: this will only work 
if both vector are of the same type): 

```{r, vectors4, exercise=TRUE}

```

```{r, vectors4-hint-1}
# Create numeric vector and assign it to the object "my_vec"
```

```{r, vectors4-hint-2}
# Perform an operation on 'my_vec' (for example x^2)
```

```{r, vectors4-hint-3}
# For fun create a character vector (call it 'my_chars') and add the values 
# of 'my_vec' to 'my_chars'. This gives an error. Why?
```

```{r, vectors4-hint-4}
# In the same vein that vectors must all be of the same class, we can only 
# perform operations on vectors of the same class.
```

</br>

### Matrices 

A matrix is a 2d vector. You probably will not use them very often, so I will 
not say much about them.

```{r, matrices, exercise=TRUE}
# Matrices = 2d vectors
matrix(1:10, nrow = 5, ncol = 2)

```

</br>


### Dataframes

Dataframes are 2d matrices with column and row names. ```IMPORTANT```: a column 
in a dataframe is like a vector, each column can be of a different type 
(factor, numeric, etc.). This is the closest equivalent to an excel sheet. You 
will probably spend most of your time in R working with dataframes.


```{r, dataframes}
mtcars <- select(mtcars, 1:4)
```

</br>

Let's look at few dataframes. The ```head()``` function shows 
the 1st six rows of a dataframe and the ```tail()``` function shows the final 
six rows. Get the first six rows of ```mtcars```

```{r, dataframes1, exercise=TRUE, exercise.setup='dataframes'}
# Get first 6 rows of mtcars

```

```{r, dataframes1-solution}
head(mtcars)
```

</br>

Now print the last 6 rows of ```USArrests```. 

```{r, dataframes2, exercise=TRUE}
# Get last 6 rows of USArrests

```

```{r, dataframes2-solution}
tail(USArrests)
```

</br>

We can get more information about a dataframe using the ```str()``` and 
```summary()``` functions. Try them below on the ```ToothGrowth``` dataset.

```{r, dataframes3, exercise=TRUE}

```

```{r, dataframes3-solution}
str(ToothGrowth)
summary(ToothGrowth)
```

