
## Manipulating dataframes with ```dplyr```

### Overview

There are some basic operations that we often have to use when working with 
dataframes, such as subsetting and creating new variables. The ```dplyr``` 
package (part of the ```tidyverse```) simplifies this process (note: it is 
always good to learn how to get the same results using base R). 


### Properties

For all of the ```dplyr``` verbs the following is true: 

- 1st arg. is a dataframe (equivalent to "data = ")
- Remaining arguments describe what to do to the dataframe
- No need for the `$` operator
- Result is a new dataframe




### Main verbs

- `select()`: return a subset of the columns of a data frame
- `filter()`: extract a subset of rows from a dataframe based on logical 
conditions
- `arrange()`: reorder rows of a dataframe
- `rename()`: rename variables in a data frame
- `mutate()`: add new variables/columns or transform existing variables
- `summarize()`: generate summary statistics of different variables in the 
dataframe


</br>





#### select()

- Use this verb to select columns of a dataframe
- You can specify a range (i.e. `columnName:columnName`)
- You can negate columns (i.e., `-mpg`)
- Special functions that only work inside `select()`
    - `starts_with(x, ignore.case = TRUE)`: names starts with `x`
    - `ends_with(x, ignore.case = TRUE)`: names ends in `x`
    - `contains(x, ignore.case = TRUE)`: selects all variables whose name 
    contains `x`
    - `matches(x, ignore.case = TRUE)`: selects all variables whose name matches 
    the regular expression `x`
    - `num_range("x", 1:5, width = 2)`: selects all variables (numerically) from 
    x01 to x05.
    - `one_of("x", "y", "z")`: selects variables provided in a character vector.
    - `everything()`: selects all variables.

**Usage example**: 

```
select(df, columnName)
select(df, -c(column1, column2))
```

</br>

**You try...**

1. Select the columns 'mpg' and 'cyl'

```{r, 'select1', exercise=TRUE}
# Select columns except 'mpg' and 'cyl'
select(mtcars, mpg, cyl) 

```

</br>

2. Select all columns except those ranging from 'mpg' to 'cyl'

```{r, 'select2', exercise=TRUE}
# Select all columns except those ranging from mpg to cyl
select(mtcars, -(mpg:cyl))  

```

</br>

3. Select all columns starting with 'c'

```{r, 'select3', exercise=TRUE}
# Select all columns starting with 'c'

```

```{r, 'select3-solution'}
select(mtcars, starts_with('c')) 
```

</br>

4. Select only the 'mpg' column

```{r, 'select4', exercise=TRUE}

```

```{r, 'select4-solution'}
# Select only the 'mpg' column
select(mtcars, mpg)
```


</br></br></br>






#### filter()

- Use ```filter()``` to do basic subsetting/indexing in a dataframe
- Use normal boolean logic
    - `==`: equal to
    - `<`: less than
    - `>`: greater than
    - `<=`: less than or equal to
    - `>=`: greater than or equal to
    - `&`: and
    - `|`: or
    - `%in%`: match variables in a set
- Important subcommand
    - `slice()`: use to select specific rows (i.e. `slice(15:30)`)

**Usage example**:

```
filter(df, bolean)
```

</br>

**You try...**

1. Filter rows in which 'cyl' equals 6 and 'disp' is greater than 100

```{r, 'filter1', exercise=TRUE}
# filter rows in which 'cyl' equals 6 and 'disp' is greater than 100

```

```{r, 'filter1-solution'}
filter(mtcars, cyl == 6 & disp > 100) 
```

</br>

2. Filter rowns in which 'cyl' is either 4 or 8

```{r, 'filter2', exercise=TRUE}
# Filter rowns in which 'cyl' is either 4 or 8

```

```{r, 'filter2-solution'}
filter(mtcars, cyl %in% c(4, 8))
```

</br>

3. Filter the rows of ```mtcars``` in which 'cyl' is equal to 8 and 'mpg' is 
less than to 15. 

```{r, 'filter3', exercise=TRUE}
# Filter rows of in which 'cyl' is equal to 8 and 'mpg' is less than 15

```

```{r, 'filter3-solution'}
filter(mtcars, cyl == 8 & mpg < 15)
```


</br></br></br>





#### arrange()

- Use ```arrange()``` if you want to order the rows of a dataframe
- Default is ascending
- To descend use `desc()`

**You try...**

1. Arrange by cyl from low to high

```{r, 'arrange1', exercise=TRUE}
# Arrange by cyl from low to high
arrange(mtcars, cyl)
```

</br>

2. Arrange by cyl from high to low

```{r, 'arrange2', exercise=TRUE}
# Arrange by cyl from high to low
arrange(mtcars, desc(cyl))
```

</br></br></br>





#### rename()

- Rename a variable (column header)
- Give the new name and then associate it with the old name

**Usage example**:

```{r, eval=FALSE, echo=TRUE}
rename(dataframe, newVar = oldVar)
```

</br>

**You try...**

1. Rename 'cyl' to 'cycle'

```{r, 'rename1', exercise=TRUE}
# Rename 'cyl' to 'cycle'
rename(mtcars, cycle = cyl)
```

</br>

2. Rename 'cyl' to 'CYL' and 'mpg' to 'miles_per_gallon'

```{r, 'rename2', exercise=TRUE}
# Rename 'cyl' to 'CYL' and 'mpg' to 'miles_per_gallon'

```

```{r, 'rename2-solution'}
# Rename 'cyl' to 'CYL' and 'mpg' to 'miles_per_gallon'
rename(mtcars, CYL = cyl, miles_per_gallon = mpg)
```


</br></br></br>






#### mutate()

- The ```mutate()``` function is used to transform existing variables
- It can also be used to reate new variables

**You try...**

```{r, mutate}
mtcars <- select(mtcars, mpg)
```

1. Center the 'mpg' variable and put it in a new column called 'mpg_c'

```{r, 'mutate1', exercise=TRUE, exercise.setup='mutate'}
# Center the 'mpg' variable and put it in a 
# new column called 'mpg_c'
mutate(mtcars, mpg_c = mpg - mean(mpg))
```

</br>

2. Standardize the 'mpg' variable and put it in a new column called 'mpg_std'


```{r, 'mutate2', exercise=TRUE, exercise.setup='mutate'}
# Standardize the 'mpg' variable and put it in a new column called 'mpg_std'

```

```{r, 'mutate2-hint-1'}
mutate(mtcars, mpg_std = (mpg - mean(mpg)) / sd(mpg))
```

```{r, 'mutate2-hint-2'}
# You can also use the ```scale()``` function. 
```

</br></br></br>







#### group_by()

- Group by categorical variables
- Useful in conjunction with `summarize()`

**You try...**

1. Calculate the mean of 'mpg' for each level of 'cyl'

```{r, 'group_by1', exercise=TRUE}
# Calculate the mean of 'mpg' for each level of 'cyl'
group_by(mtcars, cyl) %>%
  summarize(., mean_mpg = mean(mpg))
```

</br>

2. Calculate the the mean 'disp' for each level of 'vs'. 

```{r, 'group_by2', exercise=TRUE}
# Calculate the the mean 'disp' for each level of 'vs'

```

```{r, 'group_by2-solution'}
group_by(mtcars, vs) %>% 
  summarize(., mean_disp = mean(disp))
```


</br></br></br>





#### %>%

You may have noticed that the last code chunk was a little bit different. I 
used the infix operator `%>%` to link several commands together. This is called 
piping/chaining. The `%>%` operator can be read as "then". Thus, the following 
code...

```{r, 'pipe1', exercise=TRUE}
group_by(mtcars, cyl) %>%
  mutate(cMPG = mpg - mean(mpg)) %>%
  summarize(mean = mean(cMPG))
```

...can be read as: Group `mtcars` by the variable `cyl`, **then** center the 
variable `mpg` and place it in a new column, `cMPG`, **then** calculate the mean 
(for each level of `cyl`). This is extremely useful because it makes r code more 
readible and obviates the need to store a lot of variables in your computers 
memory. Let's see more examples of `%>%`.

</br>

1. Select the 'mpg', 'disp', and 'vs' columns, group by 'vs', standardize both 
variables and then calculate the mean and standard deviation (they should each 
be 0 and 1 respectively). 

```{r, 'pipe2', exercise=TRUE}
# 

```

```{r, 'pipe2-solution'}
mtcars %>%
  select(., mpg, disp, vs) %>%                    # select mpg, disp, vs
  group_by(., vs) %>%                             # group by vs
  mutate(., mpg_std = scale(mpg),                 # standardize mpg
            disp_std = scale(disp)) %>%           # standardize disp
  summarize(., mean_mpg = round(mean(mpg_std)),   # calculate mean mpg
               sd_mpg = round(sd(mpg_std)),       # calculate sd mpg
               mean_disp = round(mean(disp_std)), # calculate mean disp
               sd_disp = round(sd(disp_std)))     # calculate sd disp
```

</br></br></br>





#### Joining functions

- inner_join(x, y): gives new dataframe with columns from both x and y, but 
only includes rows that contain all columns (drops missing values)
- left_join(x, y): gives all columns of x and y, but inserts missing values for 
y (if there are no values)
- semi_join(x, y): gives all columns of x (no y added), but matches only the 
rows that match y
- anti_join(x, y): gives all columns of x (no y added), but matches only the 
rows that don't match y

</br>

|  Type |                  Action                  |
| :---- | :--------------------------------------- |
| inner | include only rows in **both** x and y    |
| left  | include all of x, and matching rows of y |
| semi  | include rows of x that match y           |
| anti  | include rows of x that **don't** match y |







