## Tidying dataframes with **tidyr**

### Overview 

This section is a continuation of [manipulating dataframes with dplyr][dplyr]. 
In this section we will work on some more basic oprations that are handy for 
tidying your data. The ```tidyr``` package (part of the ```tidyverse```) 
simplifies this process. Don't forget the ```dplyr``` verbs we already learned, 
as they will come in handy. 

So what is tidy data? **Tidy data** refers to a dataframe that has been cleaned, 
munged, modified, etc. so that it is an ideal state for data analysis (plotting, 
modeling, etc.). More specificially, this means that every column in your 
dataframe represents a variable and every row represents an observation. 
This is explained in *R for Data Science* in the following manner:

> Each variable is in a column.  
> Each observation is a row.  
> Each value is a cell.

This is also referred to as long format (as opposed to wide format). Let's look 
at an example dataset that is not tidy. How many variables are there in this 
dataset?

</br>

```{r, my_data}
my_data <- tribble(
  ~'id',    ~'spec', ~'test1', ~'test2', 
   'span01', 'g1_lo',  64.31,   69.2,
   'span02', 'g1_lo',  59.81,   63.7,
   'span03', 'g1_hi',  66.08,   70.9,
   'span04', 'g1_hi',  72.78,   79.2,
   'span05', 'g2_lo',  68.29,   75.4,
   'span06', 'g2_lo',  69.22,   76.7,
   'span07', 'g2_hi',  71.36,   77.2,
   'span08', 'g2_hi',  80.37,   88.9,
   'cata01', 'g1_lo',  75.63,   83.6,
   'cata02', 'g1_lo',  71.25,   78.8,
   'cata03', 'g1_hi',  69.09,   74.6,
   'cata04', 'g1_hi',  72.35,   80.7,
   'cata05', 'g2_lo',  71.66,   77.9,
   'cata06', 'g2_lo',  69.01,   75.0,
   'cata07', 'g2_hi',  69.86,   76.0,
   'cata08', 'g2_hi',  77.34,   85.6
) 

my_data_long <- gather(my_data, test, score, -id, -spec) 

my_data_tidy <- my_data_long %>% 
  separate(., col = spec, into = c("group", "level"), sep = "_") %>%
  separate(., col = id, into = c("lang", "id"), sep = 4)

```

```{r, print_tidy, echo=TRUE}
my_data
```

</br>

We cannot do much with the data in this format, but there are some options. For 
example, we could create a scatter plot of test1 and test2:

```{r, scatter_plot, echo=TRUE}
my_data %>% 
  ggplot(., aes(x = test1, y = test2)) + 
    geom_point()
```

We could calculate Pearson's correlation coefficient: 

```{r, corr, echo=TRUE}
cor(my_data$test1, my_data$test2)
```

Or we could compare the means using a paired samples t-test: 

```{r, ttest, echo=TRUE}
t.test(my_data$test1, my_data$test2, paired = TRUE)
```

Below we will walk through the key verbs that will allow us to tidy the data. 


### Main verbs

#### ```gather()```

- takes columns, and gathers them into key-value pairs
- this is one of the most important verbs for tidying data
- you need to tell R what names you will give to the new categorical factor and 
the new continuous variable

```{r, gather, exercise=TRUE, exercise.setup = "my_data"}
my_data_long <- gather(my_data, test, score, -id, -spec) 
my_data_long
```

Now that `test` is a factor we are able to plot the data differently. 

```{r, long_plot, exercise=TRUE, exercise.setup = "my_data"}
my_data_long %>% 
  ggplot(., aes(x = test, y = score)) + 
    geom_boxplot()
```

#### ```separate()```

- This verb is used to separate chars in a column
- essentially you create two variables from one column
- This is useful if you carefully planned your experimental outputs when you 
were collecting data (i.e. with you use the same number of chars for group 
names, common separators, etc.)

```{r, separate, exercise=TRUE, exercise.setup = "my_data"}
my_data_long %>% 
  separate(., col = spec, into = c("group", "level"), sep = "_") %>%
  separate(., col = id, into = c("lang", "id"), sep = 4)
```

#### ```spread()```

- This verb does the oposite of gather
- essentially you spread the data from long to wide
- this is useful if you already have tidy data and you need to make a scatter 
plot

```{r, spread, exercise=TRUE, exercise.setup = "my_data"}
my_data_tidy %>% 
  spread(., lang, score) 
```

```{r, spread_plot, exercise=TRUE, exercise.setup = "my_data"}
my_data_tidy %>% 
  spread(., lang, score) %>% 
  ggplot(., aes(x = cata, y = span, color = group, shape = level)) + 
    geom_point()
```

#### ```unite()``` 

- You will rarely use this verb, I think
- Use it to create a single column from two columns
- I don't know if I have ever needed to use it

```{r, unite, exercise=TRUE, exercise.setup = "my_data"}
my_data_tidy %>% 
  unite(., col = participant, lang, id, sep = "_", remove = FALSE) %>% 
  select(., -id)
```

To summarize, we can use the ```tidyr``` package to tidy our data in a way that 
makes is ideal for plotting and modeling. The functions ```gather()``` and 
```separate()``` are opposites. They can be used to convert our data from wide 
to long, and vice versa. The functions ```spread()``` and ```unite()``` are 
also complementary and can be used to split one column into two, or combine two 
columns into one. 

Continue to the next section to put all your new skills to work in a quiz. 



[dplyr]: https://jvcasill.shinyapps.io/r_for_linguists/#section-manipulating-dataframes-with-dplyr
