---
title: "Intro to R"
subtitle: "Day 6"
date: "Class date: 02/21/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/06/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```

# Libraries

```{r}
library(tidyverse)
```

# T-tests

## Independent samples

|  Name | group | score |
| :---- | :---- | :---- |
| John  | g1    | 250   |
| Jane  | g1    | 340   |
| Jimmy | g2    | 460   |
| Jessy | g2    | 200   |

Let's turn the table into a dataframe.

```{r}
name1 <- c('John', 'Jane', 'Jimmy', 'Jessy')
group1 <- c('g1', 'g1', 'g2', 'g2')
score1 <- c(250, 340, 460, 200)
ind_samp <- data.frame(name = name1, group = group1, score = score1)
print(ind_samp)
```

Let's test to see if the difference between the score means 
of 'g1' and 'g2' is equal to 0. 

```{r}
# t-test: independent 2-group
t.test(score ~ group, data = ind_samp)
```



## Paired samples

|   Name   | Test1 | Test2 |
| :------- | :---- | :---- |
| Mike     | 35    | 67    |
| Melanie  | 50    | 46    |
| Melissa  | 90    | 86    |
| Mitchell | 78    | 91    |

Let's turn the table into a dataframe.

```{r}
name2 <- c('Mike', 'Melanie', 'Melissa', 'Mitchell')
test1 <- c(35, 50, 90, 78)
test2 <- c(67, 46, 86, 91)
prd_samp <- data.frame(test1, test2)
print(prd_samp)
```

Let's test to see if the difference between 'test1' and 
'test2' is equal to 0. 

```{r}
# t-test: paired sample
t.test(test1, test2, paired = TRUE)
```

Recall that we can also use standard y-as-a-function-of-x syntax, 
but we have to convert the dataframe to long format. 

```{r}
prd_samp_long <- tidyr::gather(prd_samp, key = test, val = score)
print(prd_samp_long)

# t-test: paired sample
t.test(score ~ test, data = prd_samp_long, paired = TRUE)
```

## More examples

Run paired samples t-test on the iris dataset. 
Check the structure and pick the first 2 continuous variables.

```{r}
str(iris)
```

`Sepal.Length` and `Sepal.Width` (columns 1 and 2) are both continuous. 
Each row represents different measures of the same flower, so a paired 
test is appropriate.  

```{r}
# Paired samples t-test
t.test(iris$Sepal.Length, iris$Sepal.Width, paired = TRUE)
```

It looks like the difference between the means is not equal to 0; therefore, 
we reject the null hypothesis. Here is a plot to visualize the result.

```{r, fig.width=10, fig.retina=2}
iris %>%
  select(., Sepal.Length, Sepal.Width) %>%
  gather(., key = measure, val = value) %>%
  ggplot(., aes(x = measure, y = value, fill = measure)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot() + 
  theme_bw()
```

This is an example of how we can write up the result for publication 
in a manuscript:

>The iris data were analyzed using a paired-samples t-test, with value 
>as the dependent variable and measure (Length, Width) as a within-subjects 
>factor. The analysis revealed that the mean between the two measures 
>was not equal to 0 (t(149) = 34.82, p < 0.001). Specifically, sepal 
>length (mean: `r round(mean(iris$Sepal.Length), 2)`, sd: `r round(sd(iris$Sepal.Length), 2)`) 
>was longer than sepal width (mean: `r round(mean(iris$Sepal.Width), 2)`, 
>sd: `r round(sd(iris$Sepal.Width), 2)`) (See Figure X). The mean of 
>the differences was 2.79 cm with the 95% confidence interval falling 
>between 2.63 cm and 2.94 cm.


Notice that the write-up explains what test we used, what the variables are 
(DV, IV) and the relevant levels, the result of the test, an explanation 
of the result, including directionality of the effect.

Try to do the same thing using the `Species` factor. Let's compare the 
`Sepal.Width` of the `setosa` and `versicolor` species. How can we do this?

```{r}
str(iris)
```

`Species` is a factor with three levels. We know that a t-test can only compare 
two groups, so we need to subset the dataset in order to remove the 3rd 
level of `Species`, `virginica`. 

```{r}
# Remove 'virginica'
irisNoVirginica <- iris[iris$Species != "virginica", ]

# Independent samples t-test 
t.test(Sepal.Width ~ Species, data = irisNoVirginica, paired = FALSE)
```

Notice that I changed paired to FALSE. Why? Is each observation coming from 
the same flower? Here is a visualization of the result:

```{r, fig.width=10, fig.retina=2}
ggplot(irisNoVirginica, aes(x = Species, y = Sepal.Width, fill = Species)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot() + 
  theme_bw()
```

How would you write in prose the result of the test?

</br></br></br></br>



