---
title: "Intro to R"
subtitle: "Data structures"
date: "Class date: 01/17/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: learnr::tutorial
runtime: shiny_prerendered
---


```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE, echo=FALSE)
library(learnr)
```

# Working directories

You can set the working directory of your document using the `setwd()` 
function. You can check the current working directory using `getwd()`. 
Alternatively, you can set the working directory manually in RStudio 
(Session > Set Working Directory). 

# Install packages

Packages can be installed from RStudio using the 'Packages' tab 
(normally in the bottom right window). Search for the package you want 
and hit `enter`. You can *manually* install packages from the console 
using the `install.packages()` function. Ex. 
`install.packages('devtools')`.

# Loading packages

To load a package after you have installed it, use the `library()` 
function. Try loading the ```tidyverse``` package.

```{r, load-tidyverse, exercise=TRUE}

```

# Comments

In R, anything that comes after \# is considered a comment. Comments are 
important because they help you to remember what you did, why you did 
it, etc. Add 2 + 2 and create a relevant comment:

```{r, comment-code, exercise=TRUE}
#############################
# Always comment your code! #
#############################

# This is a comment
2 + 2
```


# R as a calculator

This isn't incredibly useful, but it gives you an idea of how R works. 
Try some simple math:

```{r, r-calculator, exercise=TRUE}

#####################
# R as a calculator #
#####################

2 + 2

4^2

(12 * 15) / 2
```




# Assigning variables

R uses objects. You must assign them using ```<-``` in order 
to store them in memory. You can them call them by typing 
the variable or calling ```print()```. Store the value '2' 
to the object 'x'

```{r, assignment, exercise=TRUE}

x <- 2
print(x)

# Perform an operation on 'x'
x + 4
```


# Data structures

```{r}

###################
# Data structures #
###################

# Vectors #

# We will create 3 vectos: x, y, and z
x <- c(1, 2, 3)
y <- c(4, 5, 6)
z <- c('a', 'b', 'c')

# We can use the print function to see the vectors in the console
print(x); print(y); print(z)

# We can perform operations on vectors
# add one to each value in the vector
# (this is much more important that it seems!)
x + 1 

# add the values in x to the values in y 
x + y 
```

```{r, eval=FALSE}
# add the values of y to the values of z (this gives an error)
# Why? What type of vector is 'x'? and 'y'?
y + z
```



```{r}
# Matrices #
# Matrices = 2d vectors

matrix(1:10, nrow = 5, ncol = 2)



# Dataframes #
# Dataframes = 2d matrices with column and row names
# IMPORTANT: a column in a dataframe is like a vector, 
# each column can be of a different type (factor, numeric, etc.)

# createa a dataframe using the vectors 'x', 'y', and 'z'
testDF <- data.frame(x, y, z)
print(testDF)

# Let's look at some more interesting dataframes
# The head() function shows the 1st six rows of a dataframe
head(mtcars)
head(USArrests)


```




















```{r, echo=FALSE}
library("htmltools")
library("bsplus")
```

# Generate a report

## Using Rmarkdown

- Open a template .Rmd file (File > New File > RMarkdown...)
- knit

## Using an R script

- .R files can be *spun* into .html files using the `spin` function  
Ex.  
`knitr::spin('myscript.R')`

<div class="row">
 <div class="col-md-2"></div>
 <div class="col-md-8">
  <div class="thumbnail">

```{r, echo=FALSE}
bs_carousel(id = "rsession") %>%
  bs_set_data(interval = FALSE) %>%
  bs_append(
            content = bs_carousel_image(src = "./assets/img/1.png"), 
            caption = bs_carousel_caption("Make sure to set the working directory")) %>%
  bs_append(
            content = bs_carousel_image(src = "./assets/img/2.png"), 
            caption = bs_carousel_caption("Use the spin function to generate report"))
```
  </div>
 </div>
 <div class="col-md-2"></div>
</div>





