---
title: "Intro to R"
subtitle: "Day 13: Working with binary data"
date: "Class date: 04/18/2017</br> Last update: `r format(Sys.time(), '%D')`"
output: 
  html_document: 
    theme: lumen
    toc: yes
    toc_depth: 2
    number_sections: yes
---

```{r, eval=FALSE, echo=FALSE, cache=FALSE}
rmarkdown::render('./programming/r/13/index.Rmd')
```

```{r, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache=FALSE)
```


First we will look at a simple case of binary data. We use 
the ```mtcars``` dataset. Let's get some descriptives.

```{r}
str(mtcars)

# make subset of mtcars
# keep column 'mpg' and 'am'

my_cars <- mtcars[, c('mpg', 'am')]

# calculate mean 'mpg' as a function of 'am'
(my_agg <- aggregate(mpg ~ am, FUN = mean, data = my_cars))
```

Now let's make a plot. We'll start simple and make a bar 
graph using the means we calculated above. 

```{r}
# make a plot
library(tidyverse)
ggplot(my_agg, aes(x = am, y = mpg)) + 
 geom_col()
```

Rather ugly, but it works. We can use ggplot to calculate 
descriptives for us. We will use ```stat_summary``` to 
calculate the mean and the standar error. This is much more 
informative than a bar graph and avoids including 0.

```{r}
ggplot(my_cars, aes(x = as.factor(am), y = mpg)) + 
  stat_summary(fun.data = 'mean_se', geom = 'errorbar') + 
  stat_summary(fun.y = 'mean', geom = 'point', color = 'red', size = 3)
```

Ok. Now let's flip our DV and IV. We will look at ```am``` as a 
function of ```mpg``` (automatic/manual transmition as a function 
of gas milleage). 

```{r}
ggplot(my_cars, aes(x = mpg, y = am)) + 
  geom_point()
```

Looks good. We can fit a model to the data using ```geom_smooth```.   

```{r}
ggplot(my_cars, aes(x = mpg, y = am)) + 
  geom_point() + 
  geom_smooth(method = 'glm', method.args = list(family = "binomial"))
```

Good. Now let's generate some data to plot something more interesting.

```{r}
# Generate data
var1  <- c(0, 0, 0, 0, 1, 1, 1)
var2  <- c(0, 0, 0, 1, 1, 1, 1)
var3  <- c(0, 0, 1, 0, 1, 1, 1)
var4  <- c(0, 0, 0, 1, 0, 1, 1)
var5  <- c(1, 0, 1, 0, 1, 1, 1)
var6  <- c(0, 0, 0, 1, 1, 1, 1)
var7  <- c(0, 0, 0, 0, 1, 1, 0)
var8  <- c(0, 0, 0, 0, 0, 1, 1)
var9  <- c(0, 1, 0, 1, 1, 0, 1)
var10 <- c(0, 0, 0, 1, 1, 1, 1)
resp <- rep(c(var1, var2, var3, var4, var5, var6, var7, var8, var9, var10), times = 10)

partN  <- 1:100
subjId <- rep(paste0('subj', partN), each = 7)
my_df  <- data.frame(subjId, 
                     stim = rep(1:7, times = 100), 
                     response = resp)

str(my_df)
head(my_df)
tail(my_df)
```

Again we can calculate descriptives and plot them. 

```{r}
(my_avg <- aggregate(response ~ as.factor(stim), FUN = mean, data = my_df))

ggplot(my_avg, aes(x = `as.factor(stim)`, y = response)) + 
  geom_col()
```

This works but let's try using what we learned above. We can also 
make the points more interpretable using ```geom_jitter```.

```{r}
ggplot(my_df, aes(x = stim, y = response)) + 
  geom_jitter(width = 0.5, height = 0.05, alpha = 0.1) + 
  geom_smooth(method = 'glm', method.args = list(family = 'binomial')) + 
  theme_bw()
```

Another option (probably a better one) is to fit an actual model and 
plot that. 

```{r}
my_glm <- glm(response ~ stim, data = my_df, family = 'binomial')
summary(my_glm)


ggplot(my_df, aes(x = stim, y = response)) + 
  stat_summary(fun.data = 'mean_se', geom = 'pointrange') + 
  stat_summary(aes(y = fitted(my_glm)), fun.y = 'mean', geom = 'line')
```

What about completely categorical data? Let's go back to the 
```mtcars``` dataset. 

```{r}
str(mtcars)

# make subset of categorical data
vs_am <- mtcars %>%
  select(vs, am)

# calculate crosstabs
(vs_am_xt <- xtabs(~ vs_am$vs + vs_am$am))

# turn the crosstab into a proportion table and plot it 
as.data.frame(prop.table(vs_am_xt, 1)) %>%
  select(., vs = vs_am.vs, am = vs_am.am, freq = Freq) %>%
  ggplot(., aes(x = vs, y = freq, fill = am)) + 
  geom_col(position = 'dodge')
```

